---
title: "Multiple subsets run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple subsets run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
This vignette demonstrates how to perform a repeated subset-based permutation test across ancestries, assess dependence across iterations, and aggregate the results using methods that account for iteration-level correlation.

# Data setup
In this example we can see a common scenario where EUR ancestry has a larger sample size than another (AFR). 
```{r data setup, fig.alt = "Bar plot showing condition imbalance across ancestries"}
library(CrossAncestryGenPhen)
library(ggplot2)

# Seed for reproducibility
seed <- 42
set.seed(seed)

# Simulate example data
p <- 100  # Number of genes
n_EUR <- 600 
n_AFR <- 40  

# Expression matrices for EUR and AFR ancestries
X <- matrix(rnorm(n_EUR * p), nrow = n_EUR, ncol = p)
Y <- matrix(rnorm(n_AFR * p), nrow = n_AFR, ncol = p)
colnames(X) <- colnames(Y) <- paste0("Gene_", seq_len(p))

# Metadata for EUR and AFR ancestries
# EUR: overrepresented compared to AFR
MX <- data.frame(
  condition = factor(c(rep("H", 400), rep("D", 200))),
  ancestry = "EUR"
)

# AFR: underrepresented compared to EUR
MY <- data.frame(
  condition = factor(c(rep("H", 10), rep("D", 30))),
  ancestry = "AFR"
)

# Visualize sample size imbalance
meta <- rbind(MX, MY)

# Plot
ggplot(meta, aes(x = ancestry, fill = condition)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Condition Imbalance Across Ancestries",
    x = "Ancestry",
    y = "Sample Count",
    fill = "Condition"
  ) 
```

# Loop run with multiple subsets
In the next step we will create the stratified subset to account for sample size but also control for the condition imbalance in the underrepresented ancestry (AFR).
```{r loop}
# Set parameters
g_col <- "condition"  
a_col <- "ancestry"  
stratify_cols <- c("condition")

# Number of subsets and permutations
n_iter <- 10
B <- 100

# Track results and sample ids
perm_results <- list()
id_log <- list()

# Loop through iterations
for (i in seq_len(n_iter)) {
  
  # Stratified ancestry sets
  split <- split_stratified_ancestry_sets(
      X = X,
      Y = Y,
      MX = MX,
      MY = MY,
      stratify_cols = stratify_cols,
      seed = seed + i
    )

  # Store sample id usage with role (train, test, inference) and iteration
  id_log[[i]] <- track_sample_ids(split, i)


  # Run permutation difference interaction analysis
  perm_res <- perm_interaction(
      X = split$test$X,
      Y = split$inference$X,
      MX = split$test$M,
      MY = split$inference$M,
      g_col = g_col,
      a_col = a_col,
      B = B,
      seed = seed + i
    )
  
  # Add up perm result
  perm_stats <- perm_res$summary_stats
  perm_stats$iteration <- i
  perm_results[[length(perm_results) + 1]] <- perm_stats
}

# Combine across iterations
perm_combined <- do.call(rbind, perm_results)
id_usage <- do.call(rbind, id_log)
```

# Output
The output will contain the combined permutation results across all iterations.
```{r output}
head(perm_combined, 10)
```

With this information we can plot the pvalue distribution.
The function `plot_pvalue_distribution()` will plot the distribution of p-values across iterations.
```{r pvalue_distribution, fig.alt = "Histogram of p-values across iterations"}
plot_pvalue_distribution(
  x = perm_combined,
  features = NULL,
  aggregation_fun = mean,
  title = "Histogram of p-values across iterations colored by mean T_obs"
)

```

# Dependence of output
Because each iteration uses the same samples from the underrepresented ancestry (AFR) and some EUR samples are shared across iteration,
there is a dependence of across iterations.
This sharing of samples can be visualized using the `plot_jaccard_heatmap()` function.
```{r jaccard_heatmap, fig.alt = "Heatmap of Jaccard index across iterations"}
plot_jaccard_heatmap(
  id_usage = id_usage,
  role = "test",
  title = "Jaccard heatmap of sample usage across test sets",
)

plot_jaccard_heatmap(
  id_usage = id_usage,
  role = "inference",
  title = "Jaccard heatmap of sample usage across inference sets",
)
```

The dependence of iterations is also visible in the correaltion of the test statistics across iterations.
This can be visualized using the `plot_correlation_heatmap()` function.
```{r correlation_heatmap, fig.alt = "Heatmap of correlation of test statistics across iterations"}
plot_correlation_heatmap(
  x = perm_combined,
  value_col = "T_obs",
  title = "Permutation T_obs correlation across subsets",
)
```

For demonstrations also the correlation of p-values can be visualized.
```{r pvalue_correlation_heatmap, fig.alt = "Heatmap of correlation of p-values across iterations"}
plot_correlation_heatmap(
  x = perm_combined,
  value_col = "p_value",
  title = "Permutation p-value correlation across subsets",
)
```

# Aggregation of iterations
## Test statistics aggregation
Ultimately, we aim to consolidate the results from multiple iterations into a single, aggregated summary.
A simple method might be to compute the mean/median of the test-statistics across iterations. 
However, this approach assumes that the iterations are independent — an assumption that often does not hold when sample splits are reused or overlap.
To properly account for this inter-iteration dependence, we use Generalized Least Squares (GLS) to aggregate the test statistics. 
GLS incorporates the correlation structure among iterations, yielding a statistically more valid and efficient estimate of the overall effect.
The function `param_gls_summary()` implements this approach under the assumption that the test statistics are approximately normally distributed.
```{r aggregation, fig.alt = "Bar plot showing aggregated p-values"}
aggregated_res <- param_gls_summary(
  x = perm_combined,
  alpha = 0.05
)

head(aggregated_res, 10)
```

However, GLS has limitations, the covariance matrix is not allowed to be singular as the inversion might fail.

## P-value aggregation