---
title: "Permutation Interaction Effect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permutation Interaction Effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
Here we showcase how permutation testing is used to assess if genotype-phenotype relationships differ across ancestries.
In omics research frameworks like `limma` are gold standard to test for differetnial expressed genes.
However, such frameworks are build on biological and statistical assuptions on the underlying data. 
In permutation testing theses assumptions are reduced, because pvalues are derived by modelling an empirical null distribution.
In case of interaction effects of the form `phenotype x ancestry`, the ancestry label is permuted creating a null, which assumes no effect by ancestry.

## Overrepresentation of EUR ancestry
In omics research EUR ancestry is often overrepresented compared to other ancestries.
Hence, data of non-Europeans is often sparse and can effect the discovery of true effects.
The framework`CrossAncestryGenPhen` tries to be fair by subsampling the overrepresented ancestry to match the sample size of the underrepresented ancestry.
The following code snippet showcases such an imbalance on a simulated dataset.

```{r data setup, fig.alt = "Bar plot showing condition imbalance across ancestries"}
library(CrossAncestryGenPhen)
library(ggplot2)

# Seed for reproducibility
seed <- 42  
set.seed(seed)

# Simulate example data
p <- 100  # Number of genes
n_EUR <- 600 
n_AFR <- 40  

# Expression matrices for EUR and AFR ancestries
X <- matrix(rnorm(n_EUR * p), nrow = n_EUR, ncol = p)
Y <- matrix(rnorm(n_AFR * p), nrow = n_AFR, ncol = p)
colnames(X) <- colnames(Y) <- paste0("Feature_", seq_len(p))

# Metadata for EUR and AFR ancestries
# EUR: overrepresented compared to AFR
MX <- data.frame(
  id = paste0("Sample_", seq_len(n_EUR)),
  condition = factor(c(rep("Control", 400), rep("Case", 200))),
  ancestry = "A"
)

# AFR: underrepresented compared to EUR
MY <- data.frame(
  id = paste0("Sample_", seq_len(n_AFR)),
  condition = factor(c(rep("Control", 10), rep("Case", 30))),
  ancestry = "B"
)

# Rownames of matrix must be smaple ids
rownames(X) <- MX$id
rownames(Y) <- MY$id

# Visualize sample size imbalance
meta <- rbind(MX, MY)

# Plot
ggplot(meta, aes(x = ancestry, fill = condition)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Condition Imbalance Across Ancestries",
    x = "Ancestry",
    y = "Sample Count",
    fill = "Condition"
  ) 
```

Running a `limma` pipeline on such imbalanced data might lead to false positive results, because the underrepresented ancestry is not well represented in the data.
One way `CrossAncestryGenPhen` tries to mitigate this is by stratifying the data.
This approach is limited to cases where the European ancestry is truly overrepreseted. 

# Single subset run of permutation interaction effect
## Stratification step
The function `split_stratified_ancestry_sets` creates a stratified subset to account for sample size but also control for the phenotype imbalance in the underrepresented ancestry.
The idea is to create a EUR `train set`, which is the remaining EUR cohort after subsampling, a EUR-subset `test set` (mimicking the underrepresented ancestry) based on the compared ancestry `inferecnce set`.

```{r stratification, fig.alt = "Bar plot showing balanced subset"}
stratify_col <- "condition" # Column to stratify on

# Split the data into stratified sets
split <- split_stratified_ancestry_sets(
  X = X,
  Y = Y,
  MX = MX,
  MY = MY,
  g_col = stratify_col,
  seed = 42
)

# Visulaize stratified sets
plot_stratified_sets(
  MX = split$test$M,
  MY = split$inference$M,
  MR = split$train$M,
  g_col = stratify_col
)
```

The output will contain `train`, `test` and `inference` sets and additional information on used strata.
Each subset contains a gene expression matrix (`$X`), a metadata frame (`$M`) and the used ids (`$ids`).

```{r split_output}
# Output
str(split)
```

The function `plot_stratified_feature` allows to plot single or multiple features per split.

```{r stratified_feature, fig.alt = "Distribution of features per split"}
plot_stratified_feature(
  X = split$test$X,
  Y = split$inference$X,
  R = split$train$X,
  MX = split$test$M,
  MY = split$inference$M,
  MR = split$train$M,
  features = NULL, # will use first 9 features
  g_col = stratify_col
)
```

## Permutation interaction analysis
The function `split_stratified_ancestry_sets` creates on downsampling of the overrepresented ancestry.
Note: A single subset only generates one estimate of the difference it is recomendet to run multiple subsets to get a more robust estimate of the interaction effect.
In case of interaction effects the `split$test` and `split$inference` sets are used.
The function `perm_interaction_effect` estimates the difference in phenotype means between the two sets. Where `X` represents one ancestry and `Y` the other ancestry.

```{r single_interaction_effect}
B <- 100  # Number of permutations

perm_res <- perm_interaction_effect(
  X = split$test$X,
  Y = split$inference$X,
  MX = split$test$M,
  MY = split$inference$M,
  g_col = stratify_col,
  B = B, 
  seed = seed
)

str(perm_res)
```

The output of the function contains three items:

1. **`summary_stats`**: Contains estimated interaction effects (`T_obs`) per feature, `p_values`, and `p_adj`, which are adjusted for multiple testing using the Benjamini–Hochberg method.

2. **`T_null`**: The observed null distribution for each feature across permutations.

3. **`B_used`**: The number of permutation iterations used to estimate the null distribution.

```{r Null_distribution, fig.alt = "Distribution of test statistic under null hypothesis"}
plot_null_distribution(
  data = perm_res, 
  p_normal = TRUE,
  p_empirical = TRUE,
  show_normal = TRUE,
  show_empirical = TRUE,
  features = NULL, # will use first 9 features
  title = "Null distribution of test statistic with two-sided p-values"
)
```

## Results of single subset run
Using the summary statistics, the effect sizes and associated p-values can be viusalized in a volcano plot using thr function `plot_volcano.R`.
```{r volcano_plot, fig.alt = "Volcano plot of interaction effect"}
plot_volcano(
  data = perm_res$summary_stats,
  x_var = "T_obs", 
  y_var = "p_param_adj", 
  sig_thr = 0.05, 
  effect_thr = 1, 
  x_label = "effect size (T_obs)",
  y_label = "-log10(p_param_adj)",
  title = "Volcano plot of interaction effect in a single subset run"
)
```


# Multiple subset run of permutation interaction effect
As previly mentioned, a single subset only generates one estimate of the difference.
To get a more robust estimate of the interaction effect, it is recommended to run multiple subsets.
In this case the pipeline is similar to run a single subset but with concerns on how to aggregate over multiple subsets.
Because the samples from the underrepresented ancestry are identical over multiple subsets this introduced dependence.
Hence, aggregating p-values or test statistics is not trivial.

## Loop run over multiple subsets

```{r multiple_subsets_run}
n_subsets <- 100  # Number of subsets to run

# Track ids and results
perm_results <- list()
id_log <- list()

for (i in 1:n_subsets) {
  
  # Stratified ancestry sets
  split <- split_stratified_ancestry_sets(
    X = X,
    Y = Y,
    MX = MX,
    MY = MY,
    g_col = stratify_col,
    seed = seed + i
  )

  # Store sample id usage with role (train, test, inference) and iteration
  id_log[[i]] <- track_sample_ids(split, i)

  perm_res <- perm_interaction_effect(
    X = split$test$X,
    Y = split$inference$X,
    MX = split$test$M,
    MY = split$inference$M,
    g_col = stratify_col,
    B = B, 
    seed = seed + i
  )

  # Add up permutation result
  perm_stats <- perm_res$summary_stats
  perm_stats$iteration <- i
  perm_results[[length(perm_results) + 1]] <- perm_stats
}

# Combine across iterations
perm_combined <- do.call(rbind, perm_results)
id_usage <- do.call(rbind, id_log)
```

The object `perm_combined` contains the combined permutation results across all subsets.
The function `plot_pvalue_distribution()` will plot the distribution of p-values across subsets.
```{r perm_combined, fig.alt = "Histogram of p-values across iterations"}
head(perm_combined, 10)

plot_pvalue_distribution(
  data = perm_combined,
  x_var = "p_param_value",
  fill_var = "T_obs",
  features = NULL, # will use first 9 features
  title = "Histogram of p-values across subsets"
)
```

## Dependence of multiple subsets
From a biological view, feautures with low p-value across subsets are more likely to be true positives.
However, from a statistical perspective, the p-values are not independent across subsets and hence to aggregate them it is important to consider.
The samples are shared across the subsets and this sharing of samples can be visualized using the `plot_jaccard_heatmap()` function.

```{r jaccard_heatmap, fig.alt = "Heatmap of Jaccard index across iterations"}
plot_jaccard_heatmap(
  data = id_usage,
  role = "test",
  title = "Jaccard heatmap of sample usage across test sets",
)

plot_jaccard_heatmap(
  data = id_usage,
  role = "inference",
  title = "Jaccard heatmap of sample usage across inference sets",
)
```

Correlation of the test statistic and the p-values can be visualized using the `plot_correlation_heatmap()` function.

```{r correlation_heatmap, fig.alt = "Heatmap of correlations"}
plot_correlation_heatmap(
  data = perm_combined,
  value_col = "T_obs",
  iter_col = "iteration",
  title = "Test statistic correlation across subsets",
)

plot_correlation_heatmap(
  data = perm_combined,
  value_col = "p_param_value",
  iter_col = "iteration",
  title = "P-value correlation across subsets",
)
```

## Aggregation across subsets
** To be continued... **