---
title: "Permutation Correlation Difference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permutation Correlation Difference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
While interaction effects are useful to test the effect of ancestry on single genes, correlation difference allows to look at a the global trend across all genes.
The same subset approach is used to correct for the overrepresentation of EUR ancestry.
This time the remaining Europeans are used as a reference or `test set` to test the correlation.
In principle the idea is that the correlation (of effect sizes of the relationship) must be higher between the remaining Europeans and the EUR-subset than between the remaining Eurpeans and non-Europeans.
Also in this case permutations are used to derive a null distribution in this case of the correlation differences between EUR-subset and the compared ancestry.
This gives a estimate of the generalizability across all genes.

## Data setup

```{r data setup, fig.alt = "Bar plot showing condition imbalance across ancestries"}
library(CrossAncestryGenPhen)
library(ggplot2)

# Seed for reproducibility
seed <- 42  
set.seed(seed)

# Simulate example data
p <- 100  # Number of genes
n_EUR <- 600 
n_AFR <- 40  

# Expression matrices for EUR and AFR ancestries
X <- matrix(rnorm(n_EUR * p), nrow = n_EUR, ncol = p)
Y <- matrix(rnorm(n_AFR * p), nrow = n_AFR, ncol = p)
colnames(X) <- colnames(Y) <- paste0("Gene_", seq_len(p))

# Metadata for EUR and AFR ancestries
# EUR: overrepresented compared to AFR
MX <- data.frame(
  id = paste0("EUR_", seq_len(n_EUR)),
  condition = factor(c(rep("Control", 400), rep("Case", 200))),
  ancestry = "EUR"
)

# AFR: underrepresented compared to EUR
MY <- data.frame(
  id = paste0("AFR_", seq_len(n_AFR)),
  condition = factor(c(rep("Control", 10), rep("Case", 30))),
  ancestry = "AFR"
)

# Rownames of matrix must be smaple ids
rownames(X) <- MX$id
rownames(Y) <- MY$id

# Visualize sample size imbalance
meta <- rbind(MX, MY)

# Plot
ggplot(meta, aes(x = ancestry, fill = condition)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Condition Imbalance Across Ancestries",
    x = "Ancestry",
    y = "Sample Count",
    fill = "Condition"
  ) 
```


# Single subset run of permutation correlation effect
The function `perm_correlation_difference()` calculates the correlation to the reference set 

```{r stratification, fig.alt = "Bar plot showing balanced subset"}
stratify_col <- "condition" # Column to stratify on

# Split the data into stratified sets
split <- split_stratified_ancestry_sets(
  X = X,
  Y = Y,
  MX = MX,
  MY = MY,
  g_col = stratify_col,
  seed = 42
)

# Visulaize stratified sets
plot_stratified_sets(
  MX = split$test$M,
  MY = split$inference$M,
  MR = split$train$M,
  g_col = stratify_col
)

# Permutation correlation difference 
B <- 100  # Number of permutations
method = "pearson" # Correlation coefficient

perm_res <- perm_correlation_difference(
  X = split$test$X,
  Y = split$inference$X,
  R = split$train$X,
  MX = split$test$M,
  MY = split$inference$M,
  MR = split$train$M,
  g_col = stratify_col,
  method = method,
  B = B, 
  seed = seed
)

str(perm_res)
```

The output of the function contains three items:

1. **`summary_stats`**: Contains estimated difference in correlation (`T_obs`) and `p_values` per features.
It also contains the correlation values to the reference `XR` and `XY`, respectively.

2. **`T_null`**: The observed null distribution for each feature across permutations.

3. **`B_used`**: The number of permutation iterations used to estimate the null distribution.
