---
title: "Single subset run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single subset run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
This tutorial shows key steps and functions in the process of generating permutation based interaction results. 

# Data setup
In this example we can see a common scenario where EUR ancestry has a larger sample size than another (AFR). 
```{r data setup, fig.alt = "Bar plot showing condition imbalance across ancestries"}
library(CrossAncestryGenPhen)
library(ggplot2)

# Seed for reproducibility
set.seed(42)

# Simulate example data
p <- 100  # Number of genes
n_EUR <- 600 
n_AFR <- 40  

# Expression matrices for EUR and AFR ancestries
X <- matrix(rnorm(n_EUR * p), nrow = n_EUR, ncol = p)
Y <- matrix(rnorm(n_AFR * p), nrow = n_AFR, ncol = p)
colnames(X) <- colnames(Y) <- paste0("Gene_", seq_len(p))

# Metadata for EUR and AFR ancestries
# EUR: overrepresented compared to AFR
MX <- data.frame(
  condition = factor(c(rep("H", 400), rep("D", 200))),
  ancestry = "EUR"
)

# AFR: underrepresented compared to EUR
MY <- data.frame(
  condition = factor(c(rep("H", 10), rep("D", 30))),
  ancestry = "AFR"
)

# Visualize sample size imbalance
meta <- rbind(MX, MY)

# Plot
ggplot(meta, aes(x = ancestry, fill = condition)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Condition Imbalance Across Ancestries",
    x = "Ancestry",
    y = "Sample Count",
    fill = "Condition"
  ) 
```

# Stratification step
In the next step we will create the stratified subset to account for imbalance in sample size but also control for the condition imbalance in the underrepresented ancestry (AFR).
```{r stratification, fig.alt = "Bar plot showing balanced subset"}
# Define column to stratify
stratify_cols <- c("condition")

# Split the data into stratified sets
split <- split_stratified_ancestry_sets(
  X = X,
  Y = Y,
  MX = MX,
  MY = MY,
  stratify_cols = stratify_cols,
  seed = 42
)

# Visulaize stratified sets
plot_stratified_sets(
  x = split,
  stratify_cols = stratify_cols
)
```

The output will contain Train, Test and Inference and info which strata were used.
For each set, the expression and metadata matrices are provided (e.g. `Matrix`: `Train$X` and `Meta`: `Train$M`).
```{r split_output}
# Output
str(split)
```

# Permutation interaction analysis
Now we can run the permutation interaction analysis on the stratified sets.
```{r interaction_analysis, fig.alt = "Bar plot showing interaction results"}
# Define the ancestry column and the condition column
g_col <- "condition"
a_col <- "ancestry"

# Define the number of permutations
B <- 1000

result <- perm_diff_interaction(
  X = split$test$X,
  Y = split$inference$X,
  MX = split$test$M,
  MY = split$inference$M,
  g_col = g_col,
  a_col = a_col,
  permute = TRUE,   # Perform permutation
  B = B,            # Number of permutations
  seed = 42,
  check_convergence = TRUE
)
```