---
title: "Single subset run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single subset run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
This tutorial shows key steps and functions in the process of generating permutation based interaction results. 

# Data setup
In this example we can see a common scenario where EUR ancestry has a larger sample size than another (AFR). 
```{r data setup, fig.alt = "Bar plot showing condition imbalance across ancestries"}
library(CrossAncestryGenPhen)
library(ggplot2)

# Seed for reproducibility
set.seed(42)

# Simulate example data
p <- 100  # Number of genes
n_EUR <- 600 
n_AFR <- 40  

# Expression matrices for EUR and AFR ancestries
X <- matrix(rnorm(n_EUR * p), nrow = n_EUR, ncol = p)
Y <- matrix(rnorm(n_AFR * p), nrow = n_AFR, ncol = p)
colnames(X) <- colnames(Y) <- paste0("Gene_", seq_len(p))

# Metadata for EUR and AFR ancestries
# EUR: overrepresented compared to AFR
MX <- data.frame(
  condition = factor(c(rep("H", 400), rep("D", 200))),
  ancestry = "EUR"
)

# AFR: underrepresented compared to EUR
MY <- data.frame(
  condition = factor(c(rep("H", 10), rep("D", 30))),
  ancestry = "AFR"
)

# Visualize sample size imbalance
meta <- rbind(MX, MY)

# Plot
ggplot(meta, aes(x = ancestry, fill = condition)) +
  geom_bar(position = "dodge", color = "black") +
  labs(
    title = "Condition Imbalance Across Ancestries",
    x = "Ancestry",
    y = "Sample Count",
    fill = "Condition"
  ) 
```

# Stratification step
In the next step we will create the stratified subset to account for sample size but also control for the condition imbalance in the underrepresented ancestry (AFR).
```{r stratification, fig.alt = "Bar plot showing balanced subset"}
# Define column to stratify
stratify_cols <- c("condition")

# Split the data into stratified sets
split <- split_stratified_ancestry_sets(
  X = X,
  Y = Y,
  MX = MX,
  MY = MY,
  stratify_cols = stratify_cols,
  seed = 42
)

# Visulaize stratified sets
plot_stratified_sets(
  x = split,
  stratify_cols = stratify_cols
)
```

The output will contain `train`, `test` and `inference` sets and additional information on used strata.
Each subset (`train`, `test`, `inference`) contains a gene expression matrix (`$X`) and a metadata frame (`$M`).
```{r split_output}
# Output
str(split)
```

# Permutation interaction analysis
Next, we use the `test` and `inference` sets to evaluate interaction effects between condition and ancestry using permutation-based inference.
The `perm_diff_interaction()` function estimates differential interactions between groups using both bootstrapping and permtuation.
```{r interaction_analysis, fig.alt = "Bar plot showing interaction results"}
# Define the ancestry column and the condition column
g_col <- "condition"
a_col <- "ancestry"

# Define the number of permutations
B <- 100

result <- perm_interaction(
  X = split$test$X,
  Y = split$inference$X,
  MX = split$test$M,
  MY = split$inference$M,
  g_col = g_col,
  a_col = a_col,
  permute = TRUE,   # Perform permutation
  B = B,            # Number of iterations (shared for both bootstrapping and permutation)
  seed = 42,
  check_convergence = TRUE
)
```

# Output
Under `summary_stats` the interaction results, including the p-values and adjusted p-values for each gene are listed.
The function uses bootstrapping to estimate the confidence intervals and SE for each interaction and permutation to calculate p-values.
It uses the same `B` for both bootstrapping and permutation, which is set to 1000 in this example.
```{r summary_stats}
head(result$summary_stats, 10)
```

More information about the statistics can be found in the `result` object.
The actual number of `B` used can be found in `result$B_used_boot` and `result$B_used_perm`, respectively.
```{r B_used}
result$B_used_boot
result$B_used_perm
```
The actual values of the test statistics can be found in `result$T_boot` and `result$T_perm`, respectively.
The distribution of the test statistics can be visualized using the `plot_T_distribution` function.
The plot shows the distribution of bootstrapped test statistics for selected genes, with dashed lines for observed values and dotted lines for 95% confidence intervals.
```{r T_boot_distribution, fig.alt = "Distribution of bootstrapped test statistics"}
plot_T_distribution(
  x = result, 
  statistic = "T_boot",
  title = "Bootstrapped distribution with 95% confidence intervals"
)
```

When having set `permute = TRUE`, the distribution of the permuted test statistics can be visualized using the `plot_T_distribution` function.
This plot displays the empirical null distribution from permutations. The dashed line indicates the observed statistic. Facet titles include empirical two-sided p-values.
```{r T_perm_distribution, fig.alt = "Distribution of permuted test statistics"}
plot_T_distribution(
  x = result, 
  statistic = "T_perm",
  title = "Permuted distribution with two-sided p-values"
)
```
